Dependent Types for Safe and Secure Web Programming

### 2014-02-03 paper/effects.tex
More paper revisions...
Tinkering and proof reading paper

\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
side-effects such as state, exceptions, and I/O as \emph{algebraic
effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
by an input and output \emph{state}, which permits effectful programs to track
the progress of a resource usage protocol. Effectful programs are written
in a monadic style, with \texttt{do}-notation, with their type stating which
specific effects are available.
Effectful
programs are described using the following data type,
in the simplest case:

@@

\subsection{Resource Protocols as Effects}

\begin{SaveVerbatim}{fileeff}
{-                                             { Input resource type }   { Output resource type } { Value } -}

>>>>

\texttt{Effects}~\cite{brady:effects} is an \idris{} library which handles
side-effects such as state, exceptions, and I/O as \emph{algebraic
effects}~\cite{Plotkin2009}. In particular, it supports parameterising effects
by an input and output \emph{state}, which permits effectful programs to track
the progress of a resource usage protocol. Effectful programs are written
in a monadic style, with \texttt{do}-notation, with their type stating which
specific effects are available:
%Effectful
%programs are described using the following data type,
%in the simplest case:

@@

\subsection{Resource Protocols as Effects}
\label{rp-effects}
\begin{SaveVerbatim}{fileeff}
{-                                             { Input resource type }   { Output resource type } { Value } -}

### 2014-02-03 paper/forms.tex
Tinkering and proof reading paper

In Sections ~\ref{formcons} and ~\ref{formhandling}, we examine implementation
of the form-handling system: namely, the effect which allows the creation of
forms, and the handling code which deserialises the data and passes it to the
user-specified handler function.

\subsection{Form Construction}
\label{formcons}
Each form element is specified to hold a particular type of data, which, assuming that the correct type of data is specified by the user, is passed directly to the handler function. In order to encapsulate this, we firstly define the allowed data types as part of an algebraic data type, \texttt{FormTy}.

@@

\noindent
Since types in \idris{} are first-class, we may use this to convert
between abstract and concrete representations of allowed form types:

@@

The general process of form construction is illustrated by the \texttt{AddTextBox}
and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure ~\ref{fig:formeffect}. These use the resource associated with the effect, \texttt{FormRes}, to construct the form. Adding a field such as a text box adds a new type,
\texttt{fty} to the list of field types, carried in the resource. When the form
is complete, the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field types, and using it to
construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

useEffects : (effs : List WebEffect) ->
             EffM m [FORM (FormRes G E)] 
                    [FORM (FormRes G effs)] ()
useEffects effs = (UseEffects effs)

@@

\noindent
We may then define handlers in a more intuitive fashion, without being
concerned with the implementation details. This allows us to write a handler
with one String argument, making use of the CGI effect, associated with the
\texttt{echo} handler function as follows:

@@

To enforce the requirement that a supplied handler function must be in the
list of available handlers, and therefore allow us to retrieve the name with
which to serialise the handler, we require a \textit{list membership proof},
\texttt{FnElem f fns}, which statically guarantees that a given item resides in
a list.

>>>>

In Sections~\ref{formcons} and~\ref{formhandling}, we examine implementation
of the form-handling system: the effect which allows the creation of
forms, and the handling code which deserialises the data and passes it to the
user-specified handler function.

\subsection{Form Construction}
\label{formcons}
Each form element holds a value of a specific type, which, assuming that the
correct type of data is given by the user, is passed directly to the
handler function. In order to encapsulate this, we firstly define the allowed
data types as part of an algebraic data type, \texttt{FormTy}.

@@

\noindent
%Since types in \idris{} are first-class, 
We convert
from an abstract to a concrete representation of an allowed form type as
follows:

@@

The general process of form construction is illustrated by the \texttt{AddTextBox}
and \texttt{Submit} operations of the \texttt{Form} effect, shown in Figure~\ref{fig:formeffect}. These use the resource associated with the effect, \texttt{FormRes}, to construct the form. Adding a field such as a text box adds a new type,
\texttt{fty} to the list of field types, carried in the resource. When the form
is complete, the \texttt{Submit} operation adds a submit button and returns the
HTML text for the form, flushing the list of field types, and using it to
construct the type for an appropriate handler function.
%
To specify a form instance, we define a function of type \texttt{UserForm}:

@@

useEffects : (effs : List WebEffect) ->
             EffM m [FORM (FormRes G E)] 
                    [FORM (FormRes G effs)] ()
useEffects effs = UseEffects effs

@@

\noindent
We may then define handlers in a more readable fashion, without being
concerned with the implementation details. This allows us to write a handler
with one String argument, making use of the CGI effect, associated with the
\texttt{echo} handler function as follows:

@@

\noindent
To enforce the requirement that a supplied handler function must be in the
list of available handlers, and therefore allow us to retrieve the name with
which to serialise the handler, we require a \textit{list membership proof},
\texttt{FnElem f fns}, which statically guarantees that a given item resides in
a list.

### 2014-02-03 paper/introduction.tex
Tinkering and proof reading paper

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL Injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

@@

%%------- EXAMPLE
\noindent
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

>>>>

Web applications, whilst ubiquitous, are also prone to incorrect construction
and security exploits such as SQL injection \cite{owasp:sqli} or cross-site
scripting \cite{owasp:xss}. Security breaches are
far-reaching, and high profile cases involve large corporations such as Sony,
who suffered a well-publicised and extremely costly SQL injection breach in
2011 \cite{ieee:sony}, and \textit{Yahoo!}, who suffered a breach in 2012
\cite{imperva:yahoo}.

@@

%%------- EXAMPLE
Let us consider a simple database access routine, written in
PHP, where we wish to obtain the name and address of every employee working in
a given department, \texttt{\$dept}. We firstly construct an object
representing a database connection, where the arguments are the database host,
user, password and name respectively:

### 2014-02-03 paper/protocols.tex
Tinkering and proof reading paper

By modelling CGI as a resource-dependent algebraic effect, we may enforce a
resource usage protocol which
prevents arbitrary IO from being performed and therefore
ensures that the headers are written correctly. We
define an effect, \texttt{Cgi}, and an associated resource,
\texttt{InitialisedCGI}, parameterised over the current state,
\texttt{CGIStep}, and containing a
\texttt{CGIInfo} record which contains information from the request. We
represent an uninitialised CGI process as the unit type, \texttt{()}.

@@

After the user code has finished execution, control returns to the library
code. At this point, the state transitions to TaskCompleted, and the headers
are written.  Finally, the headers and content are written which completes the
process. Since we parameterise the resource over a state, we may ensure that
certain operations only happen in a particular prescribed order.

@@

Here, \texttt{output} is a function which appends some output to the CGI output buffer, which is later written to the page.

@@

\begin{SaveVerbatim}{databaseeff}
{-                           { Input resource type }    { Output resource type }             { Value }           -}
data Sqlite : Effect where
  OpenDB            : DBName -> 
                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
  PrepareStatement  : QueryString -> 
                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
  BindInt           : ArgPos -> Int -> 
                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
                                                          (SQLitePSSuccess Bound))           (Maybe BindError)
  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow)         StepResult
  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow))        StepResult
  GetColumnText     : Column -> 
                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String

@@

Through the use of dependent types, we may encode these invariants statically, using the \texttt{SQLiteExecuting} resource. This is parameterised by the \texttt{ExecutionResult} data type, which indicates whether there is currently a valid row in the buffer.

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}
\noindent
Column access functions may only be called when there is a valid row in the
buffer, as signified by the input resources:

@@

\noindent
In order to provide the necessary static guarantee that there is a valid row to
process in the buffer, we make use of the \texttt{if\_valid} construct, as described in Section ~\ref{rp-effects}. The
\texttt{nextRow} function will either fetch a row into the buffer, or indicate
that a row could not be loaded, as shown by its type:

@@

\noindent
The \texttt{if\_valid} construct provides failure checking functionality, allowing different operations to be performed depending on whether or not a row was successfully fetched for processing.
%If a failure happens at any point during the computation, the resource is
%updated to reflect the failure: if, for example, the library failed to create a
%connection to the database, the resource value would be updated to
%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
%requests are made to the underlying SQLite library, in order to ensure safety.
%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
%\texttt{executeFail} utility functions allow for failures, once detected, to be
%handled by executing the appropriate sequence of state transition functions to
%dispose of any open resources and return to the initial protocol state.

>>>>

By modelling CGI as a resource-dependent algebraic effect, we may enforce a
resource usage protocol which
prevents arbitrary I/O from being performed and therefore
ensures that the headers are written correctly. We
define an effect, \texttt{Cgi}, and an associated resource,
\texttt{InitialisedCGI}, parameterised over the current state,
\texttt{CGIStep}, and containing a
\texttt{CGIInfo} record which contains information from the request. We
represent an uninitialised CGI process as the unit type, \texttt{()}.

@@

After the user code has finished execution, control returns to the library
code. At this point, the state transitions to \texttt{TaskCompleted}, 
and the headers
are written.  Finally, the headers and content are written which completes the
process. Since we parameterise the resource over a state, we may ensure that
certain operations only happen in a particular prescribed order.

@@

\noindent
Here, \texttt{output} is a function which appends some output to the CGI output buffer, which is later written to the page.

@@

\begin{SaveVerbatim}{databaseeff}
{-                           { Input resource type }    { Output resource type }             { Value }           -}

data Sqlite : Effect where
  OpenDB            : DBName -> 
                      Sqlite ()                         (Either () SQLiteConnected)          (Either SQLiteCode ())
  CloseDB           : Sqlite (SQLiteConnected)          ()                                   ()
  PrepareStatement  : QueryString -> 
                      Sqlite (SQLiteConnected)          (Either (SQLitePSFail) 
                                                          (SQLitePSSuccess Binding))         (Either SQLiteCode ())
  BindInt           : ArgPos -> Int -> 
                      Sqlite (SQLitePSSuccess Binding)  (SQLitePSSuccess Binding)            ()
  FinishBind        : Sqlite (SQLitePSSuccess Binding)  (Either SQLiteFinishBindFail
                                                          (SQLitePSSuccess Bound))           (Maybe BindError)
  ExecuteStatement  : Sqlite (SQLitePSSuccess Bound)    (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow)         StepResult
  RowStep           : Sqlite (SQLiteExecuting ValidRow) (Either (SQLiteExecuting InvalidRow)
                                                          (SQLiteExecuting ValidRow))        StepResult
  GetColumnText     : Column -> 
                      Sqlite (SQLiteExecuting ValidRow) (SQLiteExecuting ValidRow)           String

@@

We encode these invariants using the \texttt{SQLiteExecuting}
resource. This is parameterised by \texttt{ExecutionResult},
which indicates whether there is currently a valid row in the buffer:

@@

\end{SaveVerbatim}
\useverb{sqliteexecuting}
\noindent
Column access functions may only be called when there is a valid row in the
buffer, as specified by the input resources:

@@

\noindent
In order to provide the necessary static guarantee that there is a valid row to
process in the buffer, we make use of \texttt{if\_valid}, as described in Section ~\ref{rp-effects}. The
\texttt{nextRow} function will either fetch a row into the buffer, or indicate
that a row could not be loaded, as shown by its type:

@@

\noindent
The \texttt{if\_valid} construct checks for failure,
allowing different operations to be executed depending on whether or not a row
was successfully fetched for processing.
%If a failure happens at any point during the computation, the resource is
%updated to reflect the failure: if, for example, the library failed to create a
%connection to the database, the resource value would be updated to
%\texttt{OpenConn InvalidConn}. At this point, no further side-effecting
%requests are made to the underlying SQLite library, in order to ensure safety.
%The \texttt{connFail}, \texttt{stmtFail}, \texttt{bindFail} and
%\texttt{executeFail} utility functions allow for failures, once detected, to be
%handled by executing the appropriate sequence of state transition functions to
%dispose of any open resources and return to the initial protocol state.
